\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-22 pldi/draft.tex
Pass over parallelizability

To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they contain.
As mentioned before \sys focuses on parallelizing data-parallel
commands, i.e. commands that can process their input in parallel, and
encodes their characteristics by assigning them to
\emph{parallelizability classes}.
%% This information is encoded in a command's \emph{parallelizability class}.
Intuitively, a parallelizability class represents the level of synchronization required by copies of a command that execute in parallel.
It focuses on capturing characteristics that are important for a command's parallel execution, rather than the command's full observable behavior.
\sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

@@

\kk{Mention that commands change classes depending on their arguments and that we only write here (and in the table) the class of commands with no flags. Then the discussion about flags in the language part makes sense.}

% nv TODO: this is a hierarchy
Broadly, shell commands can be split into four major classes (summarized in Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on what kind of state they need when processing their input.
%% how they interact with state.
These classes are ordered in ascending difficulty of parallelization:
  later classes require more effort.
In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).

@@

\heading{Stateless Commands}
The first class, \sta, contains commands that operate on individual \todo{lines/elements}~\footnote{\todo{the choice of line as the data-quantum is deliberate, since they strike a nice balance between very coarse grained separation (files) and very fine-grained ones (individual characters). We might need to mention that in the introduction of 3.1}} of their input, without maintaining state across invocations.
These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
Stateless commands may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but when they receive empty input they return empty output.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to split inputs.
 % map :: (a -> b) -> [a] -> [b]

\kk{Proposed insertion: The choice of where how to separate data elements would change the allocation of commands in parallelization classes---} Many of these commands (about 1/3 of \sta) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
However, it turns to be  useful for cases with very large stream elements (long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3

\heading{Parallelizable Pure Commands}
The second class, \pur, contains commands that respect functional purity---\ie same outputs for same inputs---but maintain internal state across their entire pass.
The details of this state and its propagation during element processing affect their parallelizability characteristics.
Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc} simply maintains a counter.
Other commands such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.

\begin{figure}[t]
\centering
\includegraphics[width=0.20\textwidth]{\detokenize{./figs/dish_classes.pdf}}
\caption{
  \textbf{Hierarchy of parallelizability classes.}
  Parallelizability classes form a hierarchy from the easiest to parallelize (left) to the most difficult (right).
  \nv{The figure is incorrect---it's actually a lattice.}
}
\vspace{-15pt}
\label{fig:example}
\end{figure}

@@

\kk{Maybe show an example of map-fold for sort or wc}

@@

\heading{Side-effectful Commands}
The last class, \sid, are commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network. \kk{Or commands that do not produce output consuming lines of their input---thus not being amenable to data-parallelism.}
Such commands are not parallelizable without finer-grained concurrency control mechanisms that can detect side-effects across the system.

This is the largest class because it includes commands related to the file-system---a central abstractions to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.

\nv{
  A significant part of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.
  For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that is not writable by user-space scripts.
  Why aren't these parallelizable?
}

@@

\nv{Note command breadth/impossibility of our task.}
% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html

@@

\heading{Key Concerns}
\sys's annotations focused on three crucial concerns:
  (i) a command's parallelizability class,
  (ii) for commands that support multiple inputs, the characteristics of input consumption, and
  (iii) for commands with multiple flags (options), the handling of \todo{competing sets} of flags.
As the first concern was discussed extensively in the previous section, we now focus on the latter two.\footnote{
  \nv{Note to self: until now we have been talking only about parallelizability, and now we introduce two concerns that haven't been mentioned before.}
}

@@

\heading{Example Annotations}
A record contains a set of clauses, each one identified by a predicate on the commands options.
Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
% \footnote{
%   Note that the single character \ttt{-} is not an option as it is most often used to designate that a command reads from its standard input.
%   \nv{
%     There are options of of the form --f "x; y", and BSD options do not take \ttt{-} (like \ttt{ps aux}).
%     I am also not sure how important this last sentence is, as it did not answer a question I had---I expected arguments to be anything: if soneome wants to put ``pizza'' the DSL should not disallow it.
%   }
% }
The grammar of the language for specifying annotations is shown in Appendix~\ref{alg}.

For example, consider \ttt{comm}---a command that performs a join-like operation on its two inputs to identify common elements.
When \ttt{comm} is invoked without any flags, it produces a three-column output:
  the first column contains lines that are unique to the first input, the second column contains lines that are unique to the second input, and the last column contains lines that exist in both inputs.
A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output columns.
Its annotation record is shown below:
%
\begin{lstlisting}[language=sh, numbers=none]
comm {
  | -1 /\ -3 =>
    (stateless, [args[1]], [stdout])
  | -2 /\ -3 =>
    (stateless, [args[0]], [stdout])
  | _  =>
    (pure, [args[0], args[1]], [stdout])
}
\end{lstlisting}

\noindent
The first two clauses describe \ttt{comm} invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}.
In the first case, since the first and third columns are suppressed, \ttt{comm} can be considered in \sta by regarding its first file argument as a static ``configuration'' input;
  this input accompanies every replicated instance of the command executing in parallel.
The general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
There, \ttt{comm} is classified as \pur, since it needs to wait until it reads both files completely before it can output its results.
Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.

@@

\heading{Providing Combinators}
For most commands, the annotations are enough to allow parallelization: 
  commands either fall under \sta, having straightforward combiners, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{cmd}.
Unfortunately, there are commands with less trivial tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.

To support the parallelization of arbitrary commands in \pur, \sys allows command developers to provide custom $map$ and $fold$ functions.
In line with \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
  (i) $map$ in \sta and $fold$ is in \pur,
  (ii) $map$ can consume (or extend) the output of the original command and  $fold$ can consume (and combine) the results of multiple $map$ invocations, and
  (iii) their composition produces the same output with the original command.
If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their $map$ and $fold$ equivalents, enabling their parallelization.

>>>>

% To have any hope of parallelizing scripts, \sys first needs to understand the parallelization characteristics of the commands they contain.
\sys aims at parallelizing data-parallel commands, \ie commands that can process their input in parallel, and encodes their characteristics by assigning them to
\emph{parallelizability classes}.
%% This information is encoded in a command's \emph{parallelizability class}.
% Intuitively, a parallelizability 
A class represents the level of synchronization required by copies of a command executing in parallel.
It captures characteristics that are important for a command's parallel execution, rather than the command's full observable behavior.
\sys leans towards having a few coarse classes rather than many detailed ones---among other reasons, to simplify their understanding and use by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

@@

Broadly, shell commands can be split into four major classes (Tab.~\ref{tab:classes}) with respect to their parallelization characteristics, depending on what kind of state they need when processing their input.
%% how they interact with state.
These classes are ordered in ascending difficulty (or impossibility) of parallelization.
In this order, some classes can be thought as subsets of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without gaining any performance improvements).
Although commands can change classes depending on their arguments, we focus on no-flag defaults for now.

@@

\heading{Stateless Commands}
The first class, \sta, contains commands that operate on individual line elements of their input, without maintaining state across invocations.
These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{grep} filters out individual lines and \ttt{basename} removes a path prefix from a string.
Stateless commands may produce multiple elements---\eg \ttt{tr} may insert {\sc NL} tokens---but when they receive empty input they return empty output.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to split inputs.
 % map :: (a -> b) -> [a] -> [b]

The choice of line as the data element strikes a convenient balance between coarse-grained (files) and fine-grained separation (characters) while staying in line with \unix's core abstractions.
This choice can affect the allocation of commands in \sta, as many of its commands (about 1/3) are stateless \emph{within} a stream element---\eg\ttt{tr} transliterates characters within a line, one at a time---enabling further parallelization by splitting individual lines.
This feature may seem of limited use, as these commands are computationally inexpensive, precisely due to their narrow focus.
However, it turns out to be  useful for cases with very large stream elements (\ie long lines) such the \ttt{.fastq} format used in bioinformatics pipelines.  % TODO pointer to 6.3

\heading{Parallelizable Pure Commands}
The second class, \pur, contains commands that respect functional purity---\ie same outputs for same inputs---but maintain internal state across their entire pass.
The details of this state and its propagation during element processing affect their parallelizability characteristics.
Some commands are easy to parallelize, because they maintain trivial state and are commutative---\eg \ttt{wc} simply maintains a counter.
Other commands such as \ttt{sort}, maintain more complex invariants that have to be taken into account when merging partial results.
% 
% \begin{figure}[t]
% \centering
% \includegraphics[width=0.20\textwidth]{\detokenize{./figs/dish_classes.pdf}}
% \caption{
%   \textbf{Hierarchy of parallelizability classes.}
%   Parallelizability classes form a hierarchy from the easiest to parallelize (left) to the most difficult (right).
%   \nv{The figure is incorrect---it's actually a lattice.}
% }
% \vspace{-15pt}
% \label{fig:hierarchy}
% \end{figure}

@@

% \kk{Maybe show an example of map-fold for sort or wc}
% \nv{We agreed to put it end of §5}

@@

\heading{Side-effectful Commands}
The last class, \sid, are commands that have side-effects across the system---for example, updating environment variables, interacting with the filesystem, and accessing the network.
Such commands are not parallelizable without finer-grained concurrency control mechanisms that can detect side-effects across the system.

This is the largest class, for two reasons.
First, it includes commands related to the file-system---a central abstractions to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations such as access control and device driving. % is grafted upon the file-system.
Second, it contains commands that do not consume input or do not produce output---and thus are not amenable to data-parallelism.
For example, \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information that
  %A significant part of these commands (a ratio of 21:4) only \emph{read} values, and usually values that are never written by user code.

@@

% TODO: \nv{Note command breadth/impossibility of our task.}
% https://nextbreakpoint.com/posts/article-compile-code-with-docker.html

@@

\heading{Key Concerns}
\sys's annotations focused on three crucial concerns:
  (i) a command's parallelizability class,
  (ii) for commands that support multiple inputs, the characteristics of input consumption, and
  (iii) for commands with multiple flags (options), the handling of \todo{competing sets} of flags.
As the first concern was discussed extensively in the previous section, we now focus on the latter two.

@@

\heading{Example Annotations}
An annotation record contains a set of clauses, each one identified by a predicate on the command options.
Each clause maps to an assignment for the command, indicating its class and the sequence of its inputs and outputs.
Annotations treat flag arguments differently from file arguments, by checking if they start with a \ttt{-}.
% \footnote{
%   Note that the single character \ttt{-} is not an option as it is most often used to designate that a command reads from its standard input.
%   \nv{
%     There are options of of the form --f "x; y", and BSD options do not take \ttt{-} (like \ttt{ps aux}).
%     I am also not sure how important this last sentence is, as it did not answer a question I had---I expected arguments to be anything: if soneome wants to put ``pizza'' the DSL should not disallow it.
%   }
% }
The annotation grammar is shown in Appendix~\ref{alg}.

For example, consider \ttt{comm}, which performs a join-like operation on its two inputs to identify common elements.
When \ttt{comm} is invoked without any flags, it produces a three-column output:
  the first column contains lines that are unique to the first input, the second column contains lines that are unique to the second input, and the last column contains lines that exist in both inputs.
A user can invoke \ttt{comm} with any combination of flags \ttt{-1}, \ttt{-2}, or \ttt{-3} to suppress the corresponding output columns.
Its annotation record is shown below:
%
\begin{lstlisting}[language=sh, numbers=none]
comm {
| -1 /\ -3 => (S, [arg[1]], [stdout])
| -2 /\ -3 => (S, [arg[0]], [stdout])
| _  => (P, [arg[0], arg[1]], [stdout])
}
\end{lstlisting}

\noindent
The first two clauses describe \ttt{comm} invocations with options \ttt{-1} (or \ttt{-2}) and \ttt{-3}.
In the first case, since the first and third columns are suppressed, \ttt{comm} can be considered in \sta by regarding its first file argument as a static ``configuration'' input;
  this input accompanies every replicated instance of the command executing in parallel.
The general case, which includes \ttt{comm} invocations without any options, is captured by the third clause.
There, \ttt{comm} is classified as \pur as it needs to % wait until it 
  read both inputs completely before it can output its results.
Its inputs are its first and second file arguments (\ie excluding any options), and its output is produced in \ttt{stdout}.

@@

\heading{Providing Combinators}
For most commands, the annotations are enough to allow parallelization: 
  commands either fall under \sta, having straightforward combiners, or fall under a subset of \pur that can be parallelized using divide-and-conquer techniques~\sx{optimizer}.
Unfortunately, there are commands that require less obvious tactics---after all, \unix commands are Turing-complete programs and failing to plan for this generality would inhibit \sys's usefulness.

To support the parallelization of arbitrary commands in \pur, \sys allows supplying custom \emph{map} and \emph{fold} functions.
In line with the \unix philosophy, these functions can be written in any language as long as they conform to a few invariants:
  (i) \emph{map} is in \sta and \emph{fold} is in \pur,
  (ii) \emph{map} can consume (or extend) the output of the original command and  \emph{fold} can consume (and combine) the results of multiple \emph{map} invocations, and
  (iii) their composition produces the same output with the original command.
If these invariants are met, after DFG analysis~\sx{ir} \sys can replace occurrences of these commands with their \emph{map} and \emph{fold} equivalents, enabling their parallelization.
